---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# Fit regional trends to site-specific abundence data

This package fits a log-linear trend models to regions aggregated over sites. The sites may contain missing surveys that are not temporally aligned with the missing data at other sites, making direct aggregation impossible. The functions within the package model the indivdual sites with a semi-parametric model to interpolate missing data from which regional aggregations can be made. After model fitting, one can sample from the posterior predictive distribution of the regional aggregations Then calculate the log-linear trend over the time period of interest as a derived parameter. Using the posterior predictive distribution allows incorporation of both parameter uncertainty as well as uncertainty due to sampling the local abundance processes.

The model fitted to each site is given by the hierarchical specification:
$$
[y_{it}] \sim \text{Tweedie}(\mu_{it}, \phi, p)
$$



## Example

Load packages for this example
```{r}
library(agTrend.gam)
```

Now we can load the data that is included with the `agTrend.gam` package and filter it to the data we want for this example (i.e., 1990-2016). Now we'll add a photo method covariate to data (oblique photos prior to 2004 surveys = 1)

```{r}
data(wdpsNonpups)
wdpsNonpups = wdpsNonpups %>% filter(YEAR>=1990 & YEAR<=2012) %>% mutate(obl = as.integer(YEAR<2004))
```

The data is then converted to the form necessary for the site abundance imputing function 
```{r}
fit_data = wdpsNonpups %>% make.agtrend.data(abundance.name=COUNT, site.name=SITE, time.name=YEAR) 
```

Now, we count the number of positive counts at each site so that we can remove sites that had only 1 positive count
```{r}
fit_data <- fit_data %>% 
  filter(num_nonzero>1 & n_survey*avg_abund>5)
```

The next step involves creating a prior distribution list for MCMC site updating. An informative prior for the survey methodology correction is obtained from analysis of another data set. 
```{r}
data("photoCorrection")
photoCorrection %>% mutate(log_ratio = log(OBLIQUE/VERTICAL)) -> photoCorrection
gamma_0 = photoCorrection %>% summarize(mean(log_ratio)) %>% as.double()
gamma_se = photoCorrection %>% summarize(sd(log_ratio)/sqrt(n())) %>% as.double()

obs.prior=list(gamma.mean=gamma_0, gamma.vcov=gamma_se^2)
```
For the linear trend at each site, a ridge regression penalty is used such that the linear trend will be shrunk back to zero if the data do not support a trend. 


Now we begin the MCMC sampling using the `site.simulate` function. This function performs the site augmentation and samples from the posterior predictive distribution of the count data. 

```{r}
set.seed(111) 
fit_data = site.simulate(fit_data, 
                         obs.formula=~obl-1,
                         min.k = 3, obs.prior=obs.prior, 
                         fit.only=F)
```

Now we add the regional designations to `fit_data` in order to summarize the counts by desired regions

```{r}
fit_data = fit_data %>% left_join(
  wdpsNonpups %>% select(SITE, REGION, REGION) %>% distinct(), by=c("site"="SITE")
  ) %>% mutate(TOTAL="TOTAL")
```

To demonstrate we choose to provide trends for the site region designations. First, the counts are aggregated by region:

```{r}
region_data = fit_data %>% ag.abundance(REGION)
```

The the abunance can be summarized with the following function call
```{r}
region_summ = region_data %>% ag.summ()
```

Which can then be visualized with with the `ggplot2` package
```{r}
p1 = ggplot(data = region_summ %>% filter(type=="prediction")) +
  geom_path(aes(y=estimate, x=time+1990)) + 
  geom_ribbon(
    aes(ymin=ci.lower, ymax=ci.upper,x=time+1990),alpha=0.2,fill="red"
    ) + facet_wrap(~REGION, ncol=2, scales = "free_y") +
  geom_pointrange(
    aes(x=time+1990, y=estimate, ymin=ci.lower, ymax=ci.upper),
    data=region_summ %>% filter(type=="realized")
    )
print(p1)
```

Finally, we get to the main purpose of the package, to estimate log-linear trends for the aggregated abundances. This is accomplished with the function `ag.trend`

```{r}
region_trends = region_data %>% ag.trend(time.range=c(2000-1990, 2012-1990))
region_trends %>% select(-trend.sample, -trend.line)
```

The fitted trendlines can be added to the plots
```{r}
tl = region_trends %>% select(REGION, trend.line) %>% unnest()
p1 = p1 +
  geom_path(aes(y=fitted, x=time+1990), data=tl, col="blue", lwd=2) +
  xlab("Year") + ylab("Count") + theme_minimal()
p1
```

## Disclaimer

<sub>This repository is a scientific product and is not official communication of the Alaska Fisheries Science Center, the National Oceanic and Atmospheric Administration, or the United States Department of Commerce. All AFSC Marine Mammal Laboratory (AFSC-MML) GitHub project code is provided on an ‘as is’ basis and the user assumes responsibility for its use. AFSC-MML has relinquished control of the information and no longer has responsibility to protect the integrity, confidentiality, or availability of the information. Any claims against the Department of Commerce or Department of Commerce bureaus stemming from the use of this GitHub project will be governed by all applicable Federal law. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by the Department of Commerce. The Department of Commerce seal and logo, or the seal and logo of a DOC bureau, shall not be used in any manner to imply endorsement of any commercial product or activity by DOC or the United States Government.</sub>
